---
title: "ETC2420 Group 53"
authors: "Sineth Wickramsinghe, Janiru Rathnapriya, Senuth Dias, Lawrence"
output: html_document
date: "2024-09-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE, error = FALSE)
library(tidyverse)
library(broom)
library(gridExtra)
library(MASS)

# Creates a histogram and smoothed histogram given a vector of values.
dataplot <- function(x, bins = 50, colour = "blue") {
  ggplot(tibble(x = x), aes(x = x, y = after_stat(density))) +
    geom_histogram(bins = bins, colour = colour, fill = colour, alpha = 0.2) +
    geom_density(colour = colour, fill = colour, alpha = 0.2) +
    theme_bw()
}
```

```{r Reading Bank Survey Data}
bank <- read_csv("banksurvey.csv")
head(bank)
```

# **Task #1**

### Descriptive Statistics on Customer Income

```{r Descriptive Statistics}

summary(bank$income) #Provides descriptive stats of customer income 
sd(bank$income) #Standard deviation of customer income 
```
### Visualising the Customer Income ($) Data 

```{r Visualising Data}
dataplot(bank$income) + 
  xlab("Customer Income ($)") + 
  ylab("Frequency") + 
  ggtitle("Distribution of Customer Income ($)") + 
  theme(plot.title = element_text(hjust = 0.5))
```


### Fitting Customer Income to Various Models

```{r Fitting Customer Income to Varuous Models}
normal_fit <- fitdistr(bank$income, "normal")
normal_fit %>% tidy()

gamma_fit <- fitdistr(bank$income, "gamma") 
gamma_fit %>% tidy()

exponential_fit <- fitdistr(bank$income, "exponential") 
exponential_fit %>% tidy()
```

### Normal Distribution QQ Plot Customer Income ($) 

```{r}
qqnorm(bank$income, 
       main = "Normal QQ Plot Cusomter Income ($)",
       xlab = "Theoretical Quantiles (Normal Distribution)", 
       ylab = "Sample (Customer Income)")
qqline(bank$income, col = "red") 
```

### Exponential Distribution QQ Plot Customer Income ($) 


```{r}
qqplot(qexp(ppoints(length(bank$income)), rate = exponential_fit$estimate), bank$income, 
       main = "Exponential QQ Plot Cusomter Income ($)",
       xlab = "Theoretical Quantiles (Exponential Distribution)", 
       ylab = "Sample (Customer Income)")
abline(0, 1, col = "red")
```



### Gamma Distribution QQ Plot Customer Income ($) 


```{r}
qqplot(qgamma(ppoints(length(bank$income)), shape = gamma_fit$estimate["shape"], rate = gamma_fit$estimate["rate"]), bank$income, 
       main = "Gamma QQ Plot Cusomter Income ($)",
       xlab = "Theoretical Quantiles (Gamma Distribution)", 
       ylab = "Sample (Customer Income)") 

abline(0, 1, col = "red")
```



## **Task 2**

### Estimating 80th Percentile of Customer Income ($) using quantile() R function 

```{r}
normal_quantile <- qnorm(0.8, mean = normal_fit$estimate[1], sd =normal_fit$estimate[2])
exponential_quantile <- qexp(0.8, rate = exponential_fit$estimate)
gamma_quantile <- qgamma(0.8, shape = gamma_fit$estimate[1], rate = gamma_fit$estimate[2])
sample_quantile <- quantile(bank$income, probs = 0.8, type = 7)

tibble(normal_quantile, exponential_quantile, gamma_quantile, sample_quantile)
```

```{r}
B <- 5000
quant <- matrix(nrow = B, ncol = 2)

for (b in 1:B) {
  temp <- sample(bank$income, replace = TRUE)
  quant[b,] <- quantile(temp, probs = 0.8, type = 7)
}

quantile(quant[, 1], c(0.025, 0.975))
```

### *Central Limit Theorem Based Confidence Interval* 

```{r}
norm_quant <- qnorm(0.8, mean = normal_fit$estimate[1], sd =normal_fit$estimate[2])
qexp(0.8, rate = exponential_fit$estimate)
qgamma(0.8, shape = gamma_fit$estimate[1], rate = gamma_fit$estimate[2])
```
```{r}
normal_mle <- normal_fit$estimate #point estimates 
normal_se <- normal_fit$sd 
normal_ci_lower <- norm_quant + qnorm(0.025) * normal_se
normal_ci_upper <- norm_quant + qnorm(0.975) * normal_se 

tibble(parameter = c("mew", "shigma"),
       mle = norm_quant,
       se  = normal_se,
       ci_lower = normal_ci_lower,
       ci_upper = normal_ci_upper)
```

### *95% Confidence Interval: CLT Method --> Normal Distribution* 

```{r}
# Initialise values for doing the bootstrap.
B <- 5000
normal_boot <- matrix(nrow = B, ncol = 2)

# Generate the bootstrap samples and calculate statistics.
for(b in 1:B) {
  temp <- sample(bank$income, replace = TRUE)
  normal_boot[b,] <- fitdistr(temp, "normal")$estimate 
}

# Calculate CIs.
normal_boot_ci_mu    <- quantile(normal_boot[, 1], c(0.025, 0.975))
normal_boot_ci_sigma <- quantile(normal_boot[, 2], c(0.025, 0.975))

# Look at the values.
normal_boot_ci_mu
normal_boot_ci_sigma
```


## **Task 3**

## **Task 4**


## **Task 5**
